---
title: "Introduction to FSTruct"
Author: Maike Morrison
Date: 2021-04-28
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fstruct-intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warnings = FALSE,
  comment = "#>"
)

library(dplyr)

Q_example <- matrix(
  c(1, "0.001_1", 1, 1, ":", 0.428121, 0.571879, 
    2, "0.001_2", 2, 1, ":", 0.402014, 0.597986, 
    3, "0.001_3", 3, 1, ":", 0.384399, 0.615601, 
    4, "0.001_4", 4, 1, ":", 0.378551, 0.621449, 
    5, "0.001_5", 5, 1, ":", 0.373431, 0.626569, 
    6, "0.001_6", 6, 1, ":", 0.368566, 0.631434, 
    7, "0.001_7", 7, 1, ":", 0.361619, 0.638381, 
    8, "0.001_8", 8, 1, ":", 0.360798, 0.639202, 
    9, "0.001_9", 9, 1, ":", 0.358637, 0.641363, 
    10, "0.001_10", 10, 1, ":", 0.357666, 0.642334
    ),
  ncol = 7, 
  byrow = TRUE
) %>% 
  data.frame %>% 
  mutate(X6 = as.numeric(X6), 
         X7 = as.numeric(X7))

```


The FSTruct package provides an $F_{ST}$-based tool to compare variability between $Q$ matrices with rows of individual ancestry coefficients (the default output of population structure inference programs such as STRUCTURE and ADMIXTURE).

The package includes three functions:

* `Q_plot` plots Q matrices using `ggplot2`
* `Q_stat` calculates Fst/Fst^max (a  normalized measure of variability) for a Q matrix
* `Q_bootstrap` generates bootstrap replicates of one or more Q matrices along with associated statistics, including Fst/Fst^max, as well as statistical tests to compare bootstrap distributions of Fst/Fst^max . 
* `Q_simulate` generates one or more Q matrices

This package accompanies (insert paper citation here).

Population structure inference software programs such as STRUCTURE and ADMIXTURE output $Q$ matrix files which look like this:

```{r echo = FALSE}
knitr::kable(Q_example, col.names = NULL)
```

For the purposes of this function, we only care about the last $k=2$ columns--the ancestry coefficients. In a $Q$ matrix, each row represents an individual....

---------------------

## Example Analysis

```{r setup}
library(fstruct)
```
